Approximation as Differentiation

### 2025-06-09 main/approx-as-tangents.tex
Lifting.
minor edits to approx-as-tangents

\subsubsection{Stable Functions as ``Smooth Functions''}

%FIXME: use cleverref

When giving a denotational semantics for sequential programming languages, such as PCF, continuous functions are too permissive. Plotkin's Parallel OR (\autoref{ex:parallel-or}, below) is continuous but does not treat its input in a way that is consistent with a sequential implementation \cite{plotkin77lcf}. Continuous functions can explore their input in a ``random access'' way, which is incompatible with the central idea in \GPS that we should be able to identify a \emph{minimal} part of the input that leads to a part of the output.

Stability is a property that can be required of monotone functions that was invented by Berry \cite{berry86} in an attempt to capture sequentiality. This was not successful (see the $\mathrm{gustave}$ function in \autoref{ex:parallel-or}), but we will see now how it is closely related to problem of computing forward and backwards maps of approximations needed in \GPS. A textbook description of stable functions in the context of domain theory is given by \citet{amadio-curien} (Chapter 12). We start with a property that is weaker than stability, but is easier to motivate in connection with derivatives of smooth functions.

@@

  Parallel OR is famous because it is not \emph{sequential}, meaning intuitively that it cannot be implemented without running the two arguments in parallel to see if one of them returns $\mathsf{tt}$. The fact that it exists in the standard domain theoretic semantics of PCF means that this semantics is incomplete for reasoning about observational equivalence in PCF. Since Parallel OR is not cm, one might hope that cm-ness is enough to capture sequentiality, and hence potentially give a fully abstract model of PCF. However, the following ternary function $\mathbb{B}_\bot^3 \to \{\top,\bot\}$ is cm but admits no sequential implementation that fixes an order that the arguments are examined in:
  \begin{displaymath}
    \begin{array}{l@{(}l@{,~}l@{,~}l@{)~}c@{~}l}
      \mathrm{gustave}&\mathsf{tt}&\mathsf{ff}&\_&=&\top \\
      \mathrm{gustave}&\mathsf{ff}&\_&\mathsf{tt}&=&\top \\
      \mathrm{gustave}&\_&\mathsf{tt}&\mathsf{ff}&=&\top \\
      \mathrm{gustave}&\_&\_&\_&=&\bot
    \end{array}
  \end{displaymath}
  Due to the way that the cases are defined, there is no way of constructing a pair of approximations for which preservation of their meet does not hold. In terms of derivatives, this makes sense in that we are only concerned about the intensional behaviour of a function at a point and its approximations. Parallel OR has two incompatible approximation behaviours at the point $(\mathsf{tt},\mathsf{tt})$. The $\mathrm{gustave}$ function does have consistent behaviour at each approximation for each point.

%  \bob{What have hyperstable functions?}
\end{example}

@@

@@

\begin{lemma}
  Stable functions are closed under identities and composition.
\end{lemma}

@@

Stability has an alternative definition in terms of Galois connections, which will be more useful for what follows. This characterisation is due to Paul Taylor\bob{cite}.

@@

The converse to this lemma (that functions that preserve meets in their forward part have a left Galois adjoint) is not true, as was demonstrated by the non-stable function in \autoref{ex:non-stable-function}. In the case when the posets $\downset{x}$ are \emph{complete}, and $f_x$ preserves infinitary meets, then we are guaranteed a left Galois adjoint. In that example, the infinite set $\{0, 1, 2, \dots, n, n-1, \dots\}$ not including $\bot$ of approximations of $0$ does not have a greatest lower bound, so the order is not complete.

@@

>>>>

\subsubsection{Bounded Meets and Conditional Multiplicativity}
\label{sec:bounded-meets-and-cm}

When giving a denotational semantics for sequential programming languages, such as PCF, continuous functions are too permissive. Plotkin's Parallel OR (\exref{parallel-or}, below) is continuous but does not explore its input in a way consistent with a sequential implementation \cite{plotkin77lcf}. Continuous functions can explore their input in a ``random access'' way, which is incompatible with the central idea in \GPS that we should be able to identify a \emph{minimal} part of the input that leads to a part of the output.

Stability is a property that can be required of monotone functions that was invented by Berry \cite{berry86} in an attempt to capture sequentiality. This was not successful (see the $\mathrm{gustave}$ function in \exref{parallel-or}), but we will see now how it is closely related to problem of computing forward and backwards maps of approximations needed in \GPS. A textbook description of stable functions in the context of domain theory is given by \citet{amadio-curien} (Chapter 12). We start with a property that is weaker than stability, but is easier to motivate in connection with derivatives of smooth functions.

@@

  Parallel OR is famous because it is not \emph{sequential}, meaning intuitively that it cannot be implemented without running the two arguments in parallel to see if one of them returns $\mathsf{tt}$. The fact that it exists in the standard domain theoretic semantics of PCF means that this semantics is incomplete for reasoning about observational equivalence in PCF. Since Parallel OR is not cm, one might hope that cm-ness is enough to capture sequentiality, and hence potentially give a fully abstract model of PCF. However, the following ternary function $\mathbb{B}_\bot^3 \to \{\top,\bot\}$ is cm but admits no sequential implementation that fixes an order that the arguments are examined in:
  \begin{displaymath}
    \begin{array}{l@{(}l@{,~}l@{,~}l@{)~}c@{~}l}
      \mathrm{gustave}&\mathsf{tt}&\mathsf{ff}&\_&=&\top \\
      \mathrm{gustave}&\mathsf{ff}&\_&\mathsf{tt}&=&\top \\
      \mathrm{gustave}&\_&\mathsf{tt}&\mathsf{ff}&=&\top \\
      \mathrm{gustave}&\_&\_&\_&=&\bot
    \end{array}
  \end{displaymath}
  Due to the way that the cases are defined, there is no way of constructing a pair of approximations for which preservation of their meet does not hold. In terms of derivatives, this makes sense in that we are only concerned about the intensional behaviour of a function at a point and its approximations. Parallel OR has two incompatible approximation behaviours at the point $(\mathsf{tt},\mathsf{tt})$. The $\mathrm{gustave}$ function does have consistent behaviour at each approximation for each point.
\end{example}

@@

\subsubsection{Stable functions and L-posets}

@@

@@

Stability has an alternative definition in terms of Galois connections, which will be more useful for what follows. This characterisation is due to Paul Taylor\bob{cite}. We first formally define Galois connections, which we used informally in \exref{introduction-example}.

\begin{definition}[Galois connection]
  \label{def:galois-connection}
  Suppose $X$ and $Y$ are posets. A \emph{Galois connection} $f \adj g: X \to Y$ is a pair of monotone functions $f: Y \to X$ and $g: X \to Y$ satisfying $y \leq g(x) \iff f(y) \leq x$ for any $x \in X$ and $y \in Y$. We refer to $f$ as the left adjoint and $g$ as the right adjoint.
\end{definition}

@@

The converse to this lemma (that functions that preserve meets in their forward part have a left Galois adjoint) is not true, as was demonstrated by the non-stable function in \exref{non-stable-function}. In the case when the posets $\downset{x}$ are \emph{complete}, and $f_x$ preserves infinitary meets, then we are guaranteed a left Galois adjoint. In that example, the infinite set $\{0, 1, 2, \dots, n, n-1, \dots\}$ not including $\bot$ of approximations of $0$ does not have a greatest lower bound, so the order is not complete.

@@

\begin{proposition}
  L-posets and stable functions form a category. \bob{and the main point is that composition of stability witnesses looks like a chain rule!}
\end{proposition}

### 2025-06-09 main/introduction.tex
Lifting.
minor edits to approx-as-tangents

\begin{example}
  The following program is written in Haskell \cite{haskell} syntax, using a list comprehension to filter a list of pairs of labels and numbers to those numbers with a given label, and then to compute the sum of the numbers:
  \begin{displaymath}
    \begin{array}{l}
      \mathrm{query} :: \mathrm{Label} \to [(\mathrm{Label}, \mathrm{Int})] \to \mathrm{Int} \\
      \mathrm{query}\,l\,\mathit{db} = \mathrm{sum}\,[ n \mid (l',n) \leftarrow \mathit{db}, l \equiv l' ]
    \end{array}
  \end{displaymath}
  With $\mathit{db} = [(\mathsf{a}, 0), (\mathsf{b}, 1), (\mathsf{a}, 1)]$, we will have $\mathit{query}\,\mathsf{a}\,\mathit{db}$ and $\mathit{query}\,\mathsf{b}\,\mathit{db}$ both evaluating to $1$.

@@

\begin{itemize}
\item For \GPS, we assume that every value has an associated lattice of {\em approximations}. For differentiable programs, every point has an associated vector space of {\em tangents}.
\item For \GPS, every program has an associated forward approximation map that takes approximations forward from the input to the output. This map {\em preserves meets}. For differentiable programs, every program has a forward derivative that takes tangents of the input to tangents of the output. The forward derivative map is {\em linear}, so it preserves addition of tangents and the zero tangent.
\item For \GPS, every program has an associated backward approximation map that takes approximations of the output back to least approximations of the input. This map {\em preserves joins}. For differentiable programs, every program has a reverse derivative that takes tangents of the output to tangents of the input. This map is again {\em linear}.
\item For \GPS, the forward and backward approximation maps are related by being a Galois connection. For differentiable programming, the forward and reverse derivatives are related by being each others' transpose.
\end{itemize}

>>>>

\begin{example}
  \label{ex:introduction-example}
  The following program is written in Haskell \cite{haskell} syntax, using a list comprehension to filter a list of pairs of labels and numbers to those numbers with a given label, and then to compute the sum of the numbers:
  \begin{displaymath}
    \begin{array}{l}
      \mathrm{query} :: \mathrm{Label} \to [(\mathrm{Label}, \mathrm{Int})] \to \mathrm{Int} \\
      \mathrm{query}\,l\,\mathit{db} = \mathrm{sum}\,[ n \mid (l',n) \leftarrow \mathit{db}, l \equiv l' ]
    \end{array}
  \end{displaymath}
  With $\mathit{db} = [(\mathsf{a}, 0), (\mathsf{b}, 1), (\mathsf{a}, 1)]$, we will have $\mathit{query}\,\mathsf{a}\,\mathit{db}$ and $\mathit{query}\,\mathsf{b}\,\mathit{db}$ both evaluating to $1$.

@@

\bob{Have stolen this text for \autoref{sec:diff-stab-summary}}
\begin{itemize}
\item For \GPS, we assume that every value has an associated lattice of {\em approximations}. For differentiable programs, every point has an associated vector space of {\em tangents}.
\item For \GPS, every program has an associated forward approximation map that takes approximations forward from the input to the output. This map {\em preserves meets}. For differentiable programs, every program has a forward derivative that takes tangents of the input to tangents of the output. The forward derivative map is {\em linear}, so it preserves addition of tangents and the zero tangent.
\item For \GPS, every program has an associated backward approximation map that takes approximations of the output back to least approximations of the input. This map {\em preserves joins}. For differentiable programs, every program has a reverse derivative that takes tangents of the output to tangents of the input. This map is again {\em linear}.
\item For \GPS, the forward and backward approximation maps are related by being a Galois connection. For differentiable programming, the forward and reverse derivatives are related by being each others' transpose.
\end{itemize}
